---
title: "Homework 5"
format: pdf
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(mdsr)
library(dbplyr)
library(DBI)
```

```{r}
# connect to the database which lives on a remote server maintained by
#   St. Olaf's IT department

library(RMariaDB)
con <- dbConnect(
  MariaDB(), host = "mdb.stolaf.edu",
  user = "ruser", password = "ruserpass", 
  dbname = "flight_data"
)
```


## On Your Own - Adapting 164 Code - HW

These problems are based on class exercises from SDS 164, so you've already solved them in R!  Now we're going to try to duplicate those solutions in SQL (but with 2023 data instead of 2013).

```{r}
# Read in 2013 NYC flights data
library(nycflights13)
flights_nyc13 <- nycflights13::flights
planes_nyc13 <- nycflights13::planes
```


1. Summarize carriers flying to MSP by number of flights and proportion that are cancelled (assuming that a missing arrival time indicates a cancelled flight).  [This was #4 in 17_longer_pipelines.Rmd.]

```{r}
# Original solution from SDS 164
flights_nyc13 |>
  mutate(carrier = fct_collapse(carrier, "Delta +" = c("DL", "9E"), 
                                      "American +"= c("AA", "MQ"), 
                                     "United +" = c("EV", "OO", "UA"))) |>
  filter(dest == "MSP") |>   
  group_by(origin, carrier) |>
  summarize(n_flights = n(), 
            num_cancelled = sum(is.na(arr_time)),
            prop_cancelled = mean(is.na(arr_time)))
```

First duplicate the output above, then check trends in 2023 across all origins.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- is.na can be replaced with CASE WHEN ArrTime IS NULL THEN 1 ELSE 0 END or with CASE WHEN cancelled = 1 THEN 1 ELSE 0 END
- CASE WHEN can also be used replace fct_collapse
# factor collapse

```{sql, connection = con}
SELECT 
  origin,
  Reporting_Airline AS carrier,
  COUNT(*) AS num_flights,
  SUM(CASE WHEN ArrTime IS NULL THEN 1 ELSE 0 END) AS num_cancelled,
  AVG(CASE WHEN cancelled = 1 THEN 1 ELSE 0 END) AS prop_cancelled
FROM flightdata
WHERE Year = 2023 AND dest = 'MSP'  
GROUP BY origin, carrier
LIMIT 0, 20;
```


2. Plot number of flights vs. proportion cancelled for every origin-destination pair (assuming that a missing arrival time indicates a cancelled flight).  [This was #7 in 17_longer_pipelines.Rmd.]

```{r}
# Original solution from SDS 164
flights_nyc13 |>
  group_by(origin, dest) |>
  summarize(n = n(),
            prop_cancelled = mean(is.na(arr_time))) |>
  filter(prop_cancelled < 1) |>
  ggplot(aes(n, prop_cancelled)) + 
  geom_point()
```

First duplicate the plot above for 2023 data, then check trends across all origins.  Do all of the data wrangling in SQL.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- use an `sql` chunk and an `r` chunk
- include `connection = ` and `output.var = ` in your sql chunk header (this doesn't seem to work with dbGetQuery()...)

```{sql, connection = con, output.var = myflightsdata}
SELECT origin, dest,
  SUM(1) AS n,
  AVG(CASE WHEN cancelled = 1 THEN 1 ELSE 0 END) AS prop_cancelled
FROM flightdata
WHERE Year = 2023
GROUP BY origin, dest
HAVING prop_cancelled < 1;

```

# this looks like a lot

```{r}
ggplot(myflightsdata, aes(n, prop_cancelled)) +
  geom_point()
```

3. Produce a table of weighted plane age by carrier, where weights are based on number of flights per plane.  [This was #6 in 26_more_joins.Rmd.]

```{r}
# Original solution from SDS 164
flights_nyc13 |>
  left_join(planes_nyc13, join_by(tailnum)) |>
  mutate(plane_age = 2013 - year.y) |>
  group_by(carrier) |>
  summarize(unique_planes = n_distinct(tailnum),
            mean_weighted_age = mean(plane_age, na.rm =TRUE),
            sd_weighted_age = sd(plane_age, na.rm =TRUE)) |>
  arrange(mean_weighted_age)
```

First duplicate the output above for 2023, then check trends across all origins.  Do all of the data wrangling in SQL.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- you'll have to merge the flights dataset with the planes dataset
- you can use DISTINCT inside a COUNT()
- investigate SQL clauses for calculating a standard deviation
- you cannot use a derived variable inside a summary clause in SELECT

For bonus points, also merge the airlines dataset and include the name of each carrier and not just the abbreviation!


```{sql, connection = con}
SELECT 
  a.name,
  f.Reporting_Airline AS carrierID,
  COUNT(DISTINCT f.TAIL_NUMBER) AS unique_planes,
  AVG(2023 - p.year) AS mean_weighted_age,
  STDDEV_SAMP(2023 - p.year) AS sd_weighted_age
FROM flightdata AS f
JOIN planes AS p ON f.TAIL_NUMBER = p.tailnum
JOIN airlines AS a ON f.Reporting_Airline = a.carrier
GROUP BY carrier
ORDER BY mean_weighted_age ASC
LIMIT 0, 10;



```