---
title: "Homework 5"
format: pdf
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(mdsr)
library(dbplyr)
library(DBI)
```

```{r}
# connect to the database which lives on a remote server maintained by
#   St. Olaf's IT department

library(RMariaDB)
con <- dbConnect(
  MariaDB(), host = "mdb.stolaf.edu",
  user = "ruser", password = "ruserpass", 
  dbname = "flight_data"
)
```


## On Your Own - Adapting 164 Code - HW


1. Summarize carriers flying to MSP by number of flights and proportion that are cancelled (assuming that a missing arrival time indicates a cancelled flight).  [This was #4 in 17_longer_pipelines.Rmd.]

First duplicate the output above, then check trends in 2023 across all origins.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- is.na can be replaced with CASE WHEN ArrTime IS NULL THEN 1 ELSE 0 END or with CASE WHEN cancelled = 1 THEN 1 ELSE 0 END
- CASE WHEN can also be used replace fct_collapse

```{sql, connection = con}
SELECT 
  origin,
  CASE 
    WHEN Reporting_Airline IN ('DL', '9E') THEN 'Delta +' 
    WHEN Reporting_Airline IN ('AA', 'MQ') THEN 'American +' 
    ELSE 'United +' END AS carrier, 
    COUNT(*) AS num_flights,
  SUM(CASE WHEN ArrTime IS NULL THEN 1 ELSE 0 END) AS num_cancelled,
  AVG(CASE WHEN cancelled = 1 THEN 1 ELSE 0 END) AS prop_cancelled
FROM flightdata
WHERE Year = 2023 AND dest = 'MSP' AND origin IN ('LGA', 'EWR', 'JFK') 
GROUP BY origin, carrier
LIMIT 0, 20;
```


2. Plot number of flights vs. proportion cancelled for every origin-destination pair (assuming that a missing arrival time indicates a cancelled flight).  [This was #7 in 17_longer_pipelines.Rmd.]

First duplicate the plot above for 2023 data, then check trends across all origins.  Do all of the data wrangling in SQL.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- use an `sql` chunk and an `r` chunk
- include `connection = ` and `output.var = ` in your sql chunk header (this doesn't seem to work with dbGetQuery()...)

```{r}
my_flights_data <- dbGetQuery(con, '
SELECT origin, dest,
  SUM(1) AS n,
  AVG(CASE WHEN cancelled = 1 THEN 1 ELSE 0 END) AS prop_cancelled
FROM flightdata
WHERE Year = 2023
GROUP BY origin, dest
HAVING prop_cancelled < 1;
')

ggplot(my_flights_data, aes(n, prop_cancelled)) +
  geom_point()
```


3. Produce a table of weighted plane age by carrier, where weights are based on number of flights per plane.  [This was #6 in 26_more_joins.Rmd.]

First duplicate the output above for 2023, then check trends across all origins.  Do all of the data wrangling in SQL.  Here are a few hints:

- use flightdata instead of flights_nyc13
- remember that flights_nyc13 only contained 2013 and 3 NYC origin airports (EWR, JFK, LGA)
- you'll have to merge the flights dataset with the planes dataset
- you can use DISTINCT inside a COUNT()
- investigate SQL clauses for calculating a standard deviation
- you cannot use a derived variable inside a summary clause in SELECT

For bonus points, also merge the airlines dataset and include the name of each carrier and not just the abbreviation!


```{sql, connection = con}
SELECT 
  a.name,
  f.Reporting_Airline AS carrierID,
  COUNT(DISTINCT f.TAIL_NUMBER) AS unique_planes,
  AVG(2023 - p.year) AS mean_weighted_age,
  STDDEV_SAMP(2023 - p.year) AS sd_weighted_age
FROM flightdata AS f
JOIN planes AS p ON f.TAIL_NUMBER = p.tailnum
JOIN airlines AS a ON f.Reporting_Airline = a.carrier
GROUP BY carrier
ORDER BY mean_weighted_age ASC
LIMIT 0, 10;

```

