---
title: "Knowledge Quiz 2"
subtitle: "Gaby Hoover"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
---

Please answer the following questions, render a pdf, and submit via Moodle by 11 PM on Thursday, Mar 26.

Guidelines: 

- No consulting with anyone else (other than clarification questions from Professor Roith)
- You may use only materials from this class (our class webpage, links on Moodle, our 3 online textbooks, files posted to the RStudio server)
- No online searches or use of large language models like ChatGPT

Pledge: 

I pledge my honor that on this quiz I have neither given nor received assistance not explicitly approved by the professor and that I an aware of no dishonest work.

- type your name here to acknowledge the pledge: Gabrielle Hoover
- OR
- place an X here if you intentionally are not signing the pledge: _______

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(nycflights13)
```


1. Here's is a crazy list that tells you some stuff about data science.  Give code that will produce **exactly** the following outputs. (There are four parts to this question.)

```{r}
kq2_data <- list(
  DRY = c("Do not", "Repeat", "Yourself"),
  knowledge = c("Knowledge", "is", "Power"),
  nyc = flights,
  integers = 1:25,
  doubles = sqrt(1:25),
  toppings = list(sauce = c("marinara", "bbq", "creamy garlic"),
               cheese = c("mozzarella", "goat", "feta"),
               meat = c("pepperoni", "sausage", "chicken", "none"),
               veggie = c("peppers", "onions", "mushrooms", "pineapple"),
               hold = "no"),
  lostpts_to_perc = cbind(pts_lost = 0:50, 
                          perc = (50 - 0:50) / 50)
)

str(kq2_data)
```

a) 

[1] "Power"     "is"        "Knowledge"

```{r}
kq2_data[[2]][c(3, 2, 1)]
```

b) 
# A tibble: 6 × 3
# Groups:   dest [1]
  dest  carrier mean_arr_delay
  <chr> <chr>            <dbl>
1 MSP   9E                8.09
2 MSP   DL                4.04
3 MSP   EV               10.5 
4 MSP   MQ                9.56
5 MSP   OO               -2   
6 MSP   UA               -5.5 

```{r}
as.tibble(kq2_data[[3]]) |>
  filter(dest == "MSP", !is.na(arr_delay)) |>
  group_by(dest, carrier) |>
  summarize(mean_arr_delay = mean(arr_delay))
```

c) Note: It must be one string and you must subset from kq2_data for the words "bbq", "goat", "chicken", and "mushrooms" 

[1] "Give me a bbq pizza with goat cheese, chicken, and mushrooms."
```{r}
kq2_data["toppings"]

pizza <- str_c("Give me a ", kq2_data$toppings$sauce[2], " pizza with ", kq2_data$toppings$cheese[2], " cheese, ", kq2_data$toppings$meat[3], ", and ", kq2_data$toppings$veggie[3], ".")

pizza
```

d) 

[1] 10.0  0.8

```{r}
list <- kq2_data[[7]][11,]

c(list[[1]][1], list[[2]][1])
```


2. Write a function called `test_genre_popularity()` that produces a single ggplot. It should allow a user to input:

  1) a tibble
  2) two song genres
  3) and a plot option to create either a boxplot or faceted histograms
  
The title of your plot should indicate the two genres being compared. The subtitle should provide the *unique* test statistic and p-value for the specific genres being used, these values should change in the subtitle when you use different genres in the function input (Hint: the `t.test()` function will automatically calculate these numbers). Finally, your function should return an error when the user provides any plot option other than "boxplot" or "histogram". See the image KQ2_plot.png for examples of what the function should output.

```{r}
bigspotify <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')

bigspotify

# function
test_genre_popularity <- function(df, genre1, genre2, plot) {
  df <- df |> filter(playlist_genre %in% c({{ genre1 }}, {{ genre2 }}))
  t.test <- t.test(track_popularity ~ playlist_genre, data = df)

  if(plot == "boxplot") {
    df |>
      ggplot(aes(x = playlist_genre, y = track_popularity)) +
      geom_boxplot() +
      labs(title = str_c("Comparing ", {{genre1}}, " and ", {{genre2}}, " genres by popularity."),
           subtitle = str_c("Two sample t-test for difference in mean popularity scores: T = ", 
                            t.test[["statistic"]], " and p-value = ",
                            t.test[["p.value"]]))
  } else if(plot == "histogram") {
    df |>
      ggplot(aes(x = track_popularity)) +
      geom_histogram() +
      facet_wrap(~playlist_genre) +
      labs(title = str_c("Comparing ", {{genre1}}, " and ", {{genre2}}, " genres by popularity."),
           subtitle = str_c("Two sample t-test for difference in mean popularity scores: T = ", 
                            t.test[["statistic"]], " and p-value = ",
                            t.test[["p.value"]]))
  } else {
    stop("Please use boxplot or histogram as the plotting option")
  }
}

# This should work
test_genre_popularity(df = bigspotify, 
                      genre1 = "rap", 
                      genre2 = "rock", 
                      plot = "boxplot")

# This should prompt an error message that says: 
#   "Please use boxplot or histogram as the plotting option"

# test_genre_popularity(df = bigspotify, 
#                      genre1 = "rap", 
#                      genre2 = "rock", 
#                      plot = "density")


# NOTE: the only genre options are pop, rap, rock, latin, r&b, and edm
#  don't worry about an error message if one of these is not used
```

3. One of the harder concepts in statistics is the understanding of what the percentage in a confidence interval really means. Many introductory students believe that 95% confidence is a statement about the likelihood of the population parameter to be within the interval bounds. While not totally off-base, a better way to think about this is that if we take 1000 of different samples we expect around 950 of those samples to create 95% CIs that contain the true population parameter. Let's explore this.

Write a for loop that takes 10,000 samples of size 45 from a normal distribution with a true population mean of 25 and a true standard deviation of 4. Then calculate the lower bound and upper bound of a 95% confidence interval for the mean using each sample. Finally, check to see what proportion of simulated samples actually contain 25. (Spoiler - it should be close to 0.95)

Here are a couple hints and reminders:

- `rnorm(n = 20, mean = 0, sd = 1)` will generate a random sample of size 20 from a Normal distribution with mean 0 and standard deviation 1.
- The equation for a 95% confidence interval is the sample mean plus/minus 1.96 times sample standard deviation divided by the square root of sample size.

$$\bar{x} \pm 1.96 \times \frac{s}{\sqrt{n}}$$
- The mean of a vector of logicals will give you a proportion.

```{r}
# original for loop
logical <- vector("logical", 10000)
for(i in 1:10000) {
  samples <- rnorm(n = 45, mean = 25, sd = 4)
  means <- mean(samples)
  upper_bound <- mean(means) + 1.96 * (4 / sqrt(45))
  lower_bound <- mean(means) - 1.96 * (4 / sqrt(45))
  sample_log <- as.tibble(samples) |> filter(value >= 25, value < 26)
  true <- sample_log |> mutate(value = ifelse(value >= 25, TRUE, FALSE))
  logical[[i]] <- as.logical(true |> slice_head())
}

as.tibble(logical) |>
  mutate(value = ifelse(value == "NA", "0", "1")) |>
  count(value)

# proportion of samples that include 25
9896 / 10000

```


Bonus: Turn your for loop from part a) into a function with inputs for `sample_size`, `num_samples`, `true_mean`, and `true_sd`. 

