---
title: "Shiny Apps Basic"
output: html_document
date: "2026-04-21"
runtime: shiny
---

## Notes

### Shiny App within an HTML doc

  - Good for getting functionality of app
  - Good for adding text
  - Parts/Sections
    - `inputPanel()`
      - setting up the ui/user selections
      - organizing the sections/layout
      - define the changeable/interactive variables and options
      - rarely do anything with a data set
    - `renderPlot()` or `renderTable()`
      - all of the plotting/wrangling
      - use `input$___` selection
        - ___ is from the `inputPanel`
  - cheat sheet available on course website
  
  - Shiny as a stand alone app
    - use app.R file for all code
    - more flexibility with layout
    - `ui()` and `server()` sections
    - see 13a_AER_Shiny.Rmd on Prof. Roith's GitHub

### Designing a dashboard

  1. Main purpose/function (is it meant to run tests, explore data, etc?)
  
  - is it explanatory (static, non-interactive, here's the plot!)
  - or is it exploratory (interactive, clean dataset, provide pathways for exploration)
  - keep it simple: have one main purpose
    
  2. Data options
  
  - different data sets
  - filtering/wrangling options
    
  3. Secondary functionality/options/plots
  
  - smaller/supporting pieces
    
  4. Sketch a layout of your dashboard
  
  5. Additional aesthetics or tweaks
  
  6. Clear titles, labels, captions...

### Shiny App continued

  - Embed within an Rmd (above)
    - editing layout can be more challenging
      - less control
  - Standalone app
    - `app.R` file
    - see above again? ui and server sections
    - more control with layout but still challenging
    
### Flexdashboard

  - Used within Rmds but not embedded in a document (creates its own)
  - layout is a lot more straight forward
  - link on how to set up layout on the instructions doc


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

This R Markdown document is made interactive using Shiny. Unlike the more traditional workflow of creating static reports, you can now create documents that allow your readers to change the assumptions underlying your analysis and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change.  This demonstrates how a standard R plot can be made interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and `sliderInput` functions create the input widgets used to drive the plot.

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",        # for the drop down, creates an object, labels it
              choices = c(10, 20, 35, 50), selected = 20),  # gives choices for options in drop down
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:", # slider
              min = 0.2, max = 2, value = 1, step = 0.2)    # values and scale
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks), # input in the Panel section is used here
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust) 
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to prevent the R code within the chunk from rendering in the document alongside the Shiny components.

# Group Exercise I:

```{r ggplot eruptions, echo = FALSE}

inputPanel(
    selectInput("n_bins", label = "Number of bins:",        
              choices = c(5, 10, 15, 20, 25, 30, 35, 40, 45, 50), selected = 15),  
  
  sliderInput("bw_adjust2", label = "Bandwidth adjustment:", 
              min = 0.1, max = 2, value = 1, step = 0.1) 
)

renderPlot(
  ggplot(faithful, aes(x = eruptions, y = after_stat(density))) +
    geom_histogram(bins = as.numeric(input$n_bins), fill = "lightgrey", color = "black") +
    geom_density(color = "blue", adjust = input$bw_adjust2) +
    labs(x = "Duration (minutes)",
         y = "Density",
         title = "Geyser eruption duration") +
    theme_classic()
)
```

